A Domain-Specific Optimizing Compiler\\for Solving Partial Differential Equations

### 2014-03-11 paper/bare_conf.tex
Adjust table for performance results

All loops are interchangeable provided that temporaries are introduced if the nest is not perfect. For the employed storage layout, the loop permutations $ijk$ and $ikj$ are likely to maximize performance. Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required every iteration. Our experiments showed that the performance loss is greater than the gain due to accumulating increments in a register along the $i$ loop. The choice between $ijk$ and $ikj$ depends on the number of load instructions that can be hoisted out of the innermost dimension. Our compiler chooses, as outer, the loop along which the number of invariant loads is smaller so that more registers are availabe to carry out the computation of the stiffness matrix.

@@

Experiments were run on two Intel machines, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and the Phi. The \texttt{icc 2013} compiler was used, with optimization level \texttt{-O2} and with auto-vectorization enabled (\texttt{-xAVX} on the Sandy Bridge, and \texttt{TODO} on the Phi). Other optimization levels performed, in general, slightly worse than \texttt{-O2}. Our code transformations were evaluated in three real-world problems expressible with Firedrake:
\begin{itemize}
\item Helmholtz
\item Advection-Diffusion
\item Burgers
\end{itemize}
The Helmholtz code has already been shown in Figure~\ref{code:helmholtz}. For Advection-Diffusion the ``Diffusion'' equation is considered, which uses the same differential operators as Helmholtz. In the Diffusion kernel, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the stiffness matrix $A$. Burgers is a time-dependent problem (i.e. the assembly is recalculated every time step based on the result of previous iterations) employing differential operators different than those of Helmholtz and Advection-Diffusion. The assembly code, partly exposed in Figure~\ref{code:burgers}, is therefore fairly different, with a larger number of basis function matrices ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) employed.

@@

For the Helmholtz 3D problem, manual implementations based on MKL BLAS were tested on Sandy Bridge. This particular kernel can be easily reduced to a sequence of four matrix-matrix multiplications that can be computed via calls to BLAS \texttt{dgemm}. In the case of $p=4$, where the stiffness matrix is of size 35$\times$35, the computation was almost twice slower than the case in which only $licm$, data alignment and padding were used. As anticipated, both extraction of matrix-matrix multiplications from analysis of the kernel's AST and invasive re-design of the Fenics Form Compiler to explicitly expose these operations will be addressed in further work. However, these experiments justify that there's a set of problems for which turning to BLAS is not beneficial in terms of performance. It is possible that employing BLAS is useful for particularly large problems, for instance Burgers on 3D meshes with $p \geq 3$, although the loss in data locality due to re-loading matrices appearing in multiple products might inhibit the performance gain.

@@

\newcommand{\massmatrixtwodresultsnorms}{
\begin{tabularx}{1.0\textwidth}{c|XX|XX|XX|XX|}
\cline{2-9}
& \multicolumn{2}{c|}{$n_f=1$} & \multicolumn{2}{c|}{$n_f=2$} & \multicolumn{2}{c|}{$n_f=3$} & \multicolumn{2}{c|}{$n_f=4$} \\
\cline{2-9}
& A & Tensor& A & Tensor& A & Tensor& A & Tensor\\
\hline
\multicolumn{1}{|c|}{$p=1,q=1$} & $6.89e{-}17$ & $9.30e{-}17$ & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
\multicolumn{1}{|c|}{$p=1,q=2$} & $6.80e{-}17$ & $1.56e{-}16$ & $1.22e{-}16$ & $7.14e{-}17$ & $1.17e{-}17$ & $1.27e{-}17$ & $8.05e{-}18$ & $\mathbf{9.77e{-}5}$ \\
\multicolumn{1}{|c|}{$p=1,q=3$} & $8.16e{-}17$ & $8.94e{-}17$ & $6.21e{-}17$ & $\mathbf{6.77e{-}5}$ & $2.44e{-}17$ & $\mathbf{7.69e{-}4}$ & $8.61e{-}17$ & $\mathbf{5.02e{-}4}$ \\
\multicolumn{1}{|c|}{$p=1,q=4$} & $3.70e{-}16$ & $\mathbf{1.83e{-}4}$ & $7.12e{-}16$ & $\mathbf{4.11e{-}4}$ & $2.65e{-}16$ & $\mathbf{5.42e{-}4}$ & $5.03e{-}16$ & $\mathbf{4.36e{-}4}$ \\
\hline
\multicolumn{1}{|c|}{$p=2,q=1$} & $2.42e{-}16$ & $2.34e{-}16$ & $1.18e{-}16$ & $3.42e{-}17$ & $2.78e{-}17$ & $\mathbf{1.30e{-}4}$ & $1.22e{-}16$ & $\mathbf{3.06e{-}4}$ \\
\multicolumn{1}{|c|}{$p=2,q=2$} & $4.50e{-}16$ & $1.47e{-}16$ & $7.47e{-}17$ & $\mathbf{7.04e{-}5}$ & $4.05e{-}16$ & $\mathbf{6.47e{-}4}$ & - & - \\
\multicolumn{1}{|c|}{$p=2,q=3$} & $1.85e{-}16$ & $\mathbf{1.65e{-}4}$ & $7.76e{-}16$ & $\mathbf{5.57e{-}4}$ & $1.88e{-}15$ & $\mathbf{1.47e{-}3}$ & - & - \\
\multicolumn{1}{|c|}{$p=2,q=4$} & $2.22e{-}15$ & $\mathbf{2.90e{-}4}$ & $4.01e{-}15$ & $\mathbf{1.42e{-}3}$ & - & - & - & - \\
\hline
\multicolumn{1}{|c|}{$p=3,q=1$} & $6.54e{-}17$ & $5.96e{-}17$ & $8.13e{-}17$ & $\mathbf{9.70e{-}5}$ & - & - & - & - \\
\multicolumn{1}{|c|}{$p=3,q=2$} & $9.43e{-}17$ & $1.26e{-}16$ & $8.99e{-}16$ & $\mathbf{5.23e{-}4}$ & - & - & - & - \\
\multicolumn{1}{|c|}{$p=3,q=3$} & $2.27e{-}16$ & $\mathbf{2.86e{-}4}$ & $5.77e{-}15$ & $\mathbf{1.81e{-}3}$ & - & - & - & - \\
\multicolumn{1}{|c|}{$p=3,q=4$} & $2.45e{-}15$ & $\mathbf{5.31e{-}4}$ & $1.62e{-}14$ & $\mathbf{2.54e{-}3}$ & - & - & - & - \\
\hline
\end{tabularx}
}
\begin{figure*}
\massmatrixtwodresultsnorms
\end{figure*}

>>>>

All loops are interchangeable provided that temporaries are introduced if the nest is not perfect. For the employed storage layout, the loop permutations $ijk$ and $ikj$ are likely to maximize performance. Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required every iteration; experiments showed that the performance loss is greater than the gain due to accumulating increments in a register along the $i$ loop. The choice between $ijk$ and $ikj$ depends on the number of load instructions that can be hoisted out of the innermost dimension. Our compiler chooses, as outer, the loop along which the number of invariant loads is smaller so that more registers are availabe to carry out the computation of the stiffness matrix.

@@

Experiments were run on two Intel machines, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and the Phi. The \texttt{icc 2013} compiler was used, with optimization level \texttt{-O2} and with auto-vectorization enabled (\texttt{-xAVX} on the Sandy Bridge, and \texttt{TODO} on the Phi). Other optimization levels performed, in general, slightly worse than \texttt{-O2}. Our code transformations were evaluated in three real-world problems expressible with Firedrake:
\begin{itemize}
\item Helmholtz
\item Advection-Diffusion
\item Burgers
\end{itemize}
The Helmholtz code has already been shown in Figure~\ref{code:helmholtz}. For Advection-Diffusion, the ``Diffusion'' equation, which uses the same differential operators as Helmholtz, is considered. In the Diffusion kernel, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the stiffness matrix $A$. Burgers is a time-dependent problem (i.e. the assembly is recalculated every time step based on the result of previous iterations) employing differential operators different than those of Helmholtz. The assembly code, partly exposed in Figure~\ref{code:burgers}, is therefore fairly different, with a larger number of basis function matrices ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) employed.

@@

For the Helmholtz 3D problem, manual implementations based on MKL BLAS were tested on Sandy Bridge. This particular kernel can be easily reduced to a sequence of four matrix-matrix multiplications that can be computed via calls to BLAS \texttt{dgemm}. In the case of $p=4$, where the stiffness matrix is of size 35$\times$35, the computation was almost twice slower than the case in which only $licm$, data alignment and padding were used. As anticipated, extraction of matrix-matrix multiplications from analysis of the kernel's AST or re-design of the Fenics Form Compiler to explicitly expose these operations will be addressed in further work. However, these experiments justify that there's a set of problems for which turning to BLAS is not beneficial in terms of performance. It is possible that employing BLAS is useful for particularly large problems, for instance Burgers on 3D meshes with $p \geq 3$, although the loss in data locality due to re-loading matrices appearing in multiple products might limit the performance gain.

@@

\newcommand{\massmatrixtwodresultsnorms}{
\begin{tabularx}{1.0\textwidth}{cccc|cc|cc|}
\cline{5-8}
& & & & \multicolumn{2}{c|}{$split$} & \multicolumn{2}{c|}{$op$-$vect$} \\
\cline{2-8}
& \multicolumn{1}{|c|}{original} & \multicolumn{1}{|c|}{$licm$} & $licm$+$ap$ & Best & Cost Model & Best & Cost Model\\
\cline{1-8}
\multicolumn{1}{|c|}{Helmholtz$\_$tri$\_$p1} & & \multicolumn{1}{|c|}{$5.49e{-}17$} & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
\cline{1-8}
\end{tabularx}
}
\begin{figure*}
\massmatrixtwodresultsnorms
\end{figure*}
